``Interestingness'' in Mathematical Theory Exploration

### 2019-10-16 PhDThesis/litreview_staticanalysis.tex
Bulk out static analysis

Some of the simplest code improvement tools are known as \emph{linters}, after
the \textsc{Lint} tool for the C programming language~\cite{Johnson78lint}.
Linters are characterised by parsing the given code and looking for occurrences
of certain syntax patterns, for example definitions which are never used, or
``no-op'' constructs (e.g. \texttt{(x~<~0)} when \texttt{x} is of
unsigned/non-negative type, making this comparison always false).

@@

At the other end of the code improvement spectrum we find
\emph{superoptimisers}~\cite{massalin1987superoptimizer}, which can take an
arbitrarily long time to find improvements, but whose deductive power is only
limited by the undecidability of the source language. Superoptimisers, such as
\textsc{Souper}~\cite{sasnauskas2017souper} and \textsc{Aha!} (``A Hacker's
Assistant'', developed to accompany \emph{Hacker's
  Delight}~\cite{warren2013hacker}) treat the given code as a specification, and
perform an open-ended search through the space of all programs to find the
optimal implementation according to some quality measure or cost model (often a
branch- and effect-free subset of machine code is used, and measured by counting
instructions or cycles).  Superoptimisers need some method to check whether a
synthesised program implements the required specification, such as a test suite
and/or a theorem prover.

Proving correctness of machine code programs is helped by the uniform, finite
nature of their domain (e.g. a fixed set of instructions, operating on a fixed
set of fixed-size registers). Higher level languages like Haskell are more
difficult to reason about and measure, since their behaviour and efficiency may
depend on unknown runtime parameters. For example, a higher-order function $f$
accepting another function $g$ as an argument may choose to memoise calls to $g$
in order to avoid evaluating identical calls multiple times; however, storing
previous results requires more memory, and each input must be looked up in the
memo table to see if it's been used before. Whether this is an optimisation or
pessimisation depends on the runtime data at each call site, including the
function argument (whether it's cheaper than performing a comparison) and the
data it will be applied to (whether it actually contains any repeated values).
In addition to this uncertainty, each function's input domain may be infinite
(such as the natural numbers), and programs may use abstract interfaces and
datatypes with a variety of possible implementations (such as lists or key/value
mappings). All these factors makes effective testing more difficult, and proofs
potentially undecidable.

We can reduce the amount of unknown parameters in a concrete program by
propagating statically-known values, for example using
\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}; however this can
dramatically increase code size, as each function may be duplicated into
multiple specialised instances. Related techniques, like \emph{symbolic
  execution} (a form of \emph{abstract interpretation}), allow approximate
execution of code, where unknown runtime data is represented as opaque
metavariables. Property- and model-checkers are similar, except they work
empirically by performing multiple runs, instantiating metavariables with
synthetic data (\quickcheck{} does this randomly, \smallcheck{} enumerates
values, etc.).

Efficiency can be calculated using techniques like cost
semantics~\cite{danner2015denotational}, with constant factors determined
through empirical testing. This cannot solve the problem of optimality depending
on the characteristics of runtime data, but it can expose these dependencies as
parameters allowing the user to select between pareto-optimal strategies.

@@

The open-ended search process of such superoptimisers is similar to application
of mathematical theory exploration to software property discovery, especially
when exploration is limited to equality properties (which is the case for
\isacosy{} and \quickspec{}). Quality metrics like efficiency (e.g.
relative instruction count) can be considered a subset of the interestingness
criteria of theory exploration. The \emph{aims} of these analyses are in
opposition, since theory exploration \emph{builds up} structure and abstraction
latent in a set of definitions, whilst superoptimisation \emph{tears down} such
layers to find minimal implementations; yet the practical necessity of
superoptimising only small snippets at a time (e.g. to find ``peephole
optimisations''~\cite{Bansal.Aiken:2006}) results in remarkably similar
discovery processes.

% TODO: Evolutionary patches
% TODO: Hutter search
% TODO: Goedel machines

>>>>

Some of the simplest code improvement tools are known as \emph{linters}, after
the \textsc{Lint} tool for the C programming language~\cite{Johnson78lint}.
Linters are standalone tools, characterised by parsing the given code and
looking for occurrences of certain syntax patterns; for example definitions
which are never used, ``no-op'' constructs (e.g. \texttt{if~(x~<~0)} when
\texttt{x} is of unsigned/non-negative type) and expressions which are valid but
usually used erroneously (e.g. \texttt{if (x = 1)} which performs assignment
rather than comparison).

@@

At the other end of the code improvement spectrum we find
\emph{superoptimisers}~\cite{massalin1987superoptimizer}, which can take an
arbitrarily long time to find improvements, but whose deductive power is only
limited by the undecidability of the source language. Superoptimisers, such as
\textsc{Souper}~\cite{sasnauskas2017souper} and \textsc{Aha!} (``A Hacker's
Assistant'', developed to accompany \emph{Hacker's
  Delight}~\cite{warren2013hacker}) treat the given code as a specification, and
perform an open-ended search through the space of all programs to find the
optimal implementation according to some quality measure or cost model.

Supercompilers are often restricted to a branch- and effect-free subset of
machine code, where cost can be reasonably estimated by counting instructions or
cycles. Higher-level languages require more elaborate estimation methods, such
as cost semantics~\cite{danner2015denotational}: such models will still contain
many parameters, some of which may be determined by empirical measurement,
whilst others may depend on characteristics of the input data. Despite this
drawback, pareto-optimal optimisations can still be identified, and either
presented to the user or tested against a realistic data set (known as
\emph{profile-guided optimisation}~\cite{TODO}).

Superoptimisers also need some method to check whether a synthesised program
implements the required specification, such as a test suite and/or a theorem
prover. Proving correctness of machine code fragments is helped by the uniform,
finite nature of their domain (a fixed set of instructions, operating on a fixed
set of fixed-size registers). Higher level languages like Haskell are more
difficult to reason about and measure, since their behaviour and efficiency may
depend on unknown runtime parameters. In particular, optimising higher-order
functions is difficult in isolation, since many performance tradeoffs cannot be
decided without more knowledge of the functions involved. For example, a cache
of previous return values (known as a \emph{memo table}) can avoid re-computing
repeated calls to a function; but whether this actually saves time, or is worth
the memory penalty, depends on how long that function takes to calculate; and
how often it will be called with same arguments.

In addition to this uncertainty, each function's input domain may be infinite
(such as the natural numbers), and programs may use abstract interfaces and
datatypes with a variety of possible implementations (such as lists or key/value
mappings). All these factors makes effective testing more difficult, and proofs
potentially undecidable.

@@

Rather than optimising individual units of code in isolation, we can instead
work ``top down'' and perform \emph{whole program optimisation}.\iffalse TODO: Cite MLTon, Stalin,
maybe others \fi In particular, for every function we may want to optimise, we
know \emph{all} call sites of that function within the program, and hence avoid
the need for a general solution. Statically-known values, including arguments to
(higher-order) functions, can be propagated through the code using techniques
like \emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}, and constant
expressions can be ahead of time; however this can dramatically increase code
size, as each function may be duplicated into multiple specialised instances.

Related techniques, like \emph{symbolic execution} (a form of \emph{abstract
  interpretation}), allow information to be statically deduced by tracing the
behaviour of code using an alternative execution model, where unknown runtime
data are represented as opaque metavariables. Property- and model-checkers are
similar, except they work dynamically by performing multiple runs, instantiating
metavariables with exhaustive or synthetic data (\quickcheck{} does this
randomly, \smallcheck{} enumerates values, etc.).

The most common code-improvement technique performed \emph{during execution} is
Just In Time (JIT) compilation~\iffalse TODO: Cite \fi, which allows an
interpreter to compile sections of code into a faster form (usually machine
code). Since this compilation takes time, it is usually reserved for sections
which are executed often enough to exceed some predefined threshold (e.g. 100).
By focusing on such ``hot paths'', the up-front costs of compiling everything
ahead of time are avoided; although compiled forms are kept only in memory,
getting recompiled as needed on each program run. Compiling at runtime also
allows specialisation with dynamically-known information; for example, inlining
specific implementations of a procedure rather than performing dynamic dispatch
(this is the problem higher-order functions discussed above). To avoid altering
semantics, such dynamic specialisations must be accompanied by a conservative
validity check (e.g. that the given arguments should use the procedure
implementation we have compiled-in), falling back to the interpreter upon
failure.

A more speculative use of optimisation during runtime has been proposed by
Hutter~\iffalse TODO: Cite \fi. As in a JIT-enabled interpreter, Hutter's
algorithm begins by directly executing the given (unoptimised) program $p$ on
the relevant input $x$ (this may be provided up front, or piecemeal throughout
execution, e.g. as a stream of user interaction events). The variables $p_{fast}$
and $t_{fast}$ are initialised to $p$ and $\infinity$ respectively, to indicate
that $p$ is the fastest implementation we have so far, and we don't yet have an
upper bound on its running time. Concurrently with the execution of $p$, two
other processes are also started:

\begin{itemize}
\item Process A searches for proofs (in a pre-specified formal system) that some
  program $p\prime$ is both semantically equivalent to $p$ and has a running
  time bounded by some other program $t$, i.e. for all inputs $i$, the
  time taken to compute $p\prime(i)$ is bounded by $t(i)$. A list is built up of
  the $(p\prime, t)$ pairs found so far.
\item Process B computes $t(x)$ for all of the bounds $t$ found so far by A.
  Computations are interleaved via a schedule similar to Levin's universal
  search algorithm~\iffalse TODO: Cite \fi, to prevent getting stuck on
  non-halting programs. Whenever a calculation of some $t(x)$ finishes, it is
  compared to $t_{fast}$: if smaller, $t_{fast}$ is set to $t(x)$, and $p_{fast}$ is
  set to $p\prime$ (the program paired with this $t$).
\end{itemize}

The execution of $p_{fast}(x)$ (initially $p(x)$) is restarted at
exponentially-spaced intervals: if the value of $p_{fast}$ gets updated,
execution will switch to the new, provably-faster program at the next restart.
Although wasteful, such restarting will only slow down execution by at most a
factor of 2; whilst switching to an updated $p_{fast}$ may bring arbitrarily large
speed-ups (depending on the problem). Since the proof search (process A) is
independent of the input $x$, it only contributes a constant factor to the
running time; although in the general case this factor will be exponential in
the length of the proof.

Hutter's algorithm is mostly of theoretical interest, being described as
``another $O()$ warning showing, how important [to running time] factors, and
even subdominant additive terms, are.''~\cite{TODO: Hutter, page 4}. Despite
this caveat, the general framework is still useful to compare with the code
improvement techniques described above. In particular, the use of open-ended
search (process A) is similar to that of superoptimisation; although the former
enumerates equivalence proofs and extracts programs from them, whilst the latter
enumerates programs and attempts to prove (or approximately determine)
equivalence. Likewise the speed estimation of process B is analogous to those
employed by superoptimisation. JIT compilation can also be achieved within this
framework, for example by including our desired optimisations in the axioms of
the formal system (making proof search trivial), biasing the search algorithm to
consider applications of these optimisations first, and including a simple cost
model for the language. Switching to faster programs can also be made
incremental, e.g. allowing (stateless) functions to be replaced individually
without having to restart execution, just like in a JIT-enabled interpreter.

The open-ended search through program and proof space performed by
superoptimisation and Hutter's algorithm are similar to that of Theory
Exploration (e.g. the discovery of equivalent expressions by \isacosy{} and
\quickspec{}). One major difference is in \emph{end result} of these techniques:
the latter tends to \emph{build up} structure and abstraction latent in a set of
definitions, whilst the former tend to \emph{tear down} such layers to find
minimal implementations.

Another seeming contrast is that Theory Exploration has no ``target'', unlike
optimisers. However, the practical necessity of limiting superoptimisers to only
small snippets of code at a time has resulted in applications like finding
``peephole optimisations''~\cite{Bansal.Aiken:2006}, whose scattershot nature is
more akin to Theory Exploration than traditional compiler optimisation.

In any case, efficiency (whether instruction count, small upper bounds, cost
semantics, etc.) can be considered part of the interestingness criteria of
Theory Exploration, and there are certainly opportunities to cross-pollinate
techniques between these fields.

% TODO: Evolutionary patches
% TODO: Goedel machines
